---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# **`verywise`**
## vertex-wise, whole-brain linear mixed models

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of `verywise` is to offer a flexible, user-friendly interface to whole-brain analysis of neuro-imaging data that has been pre-processed using [FreeSurfer](https://surfer.nmr.mgh.harvard.edu/). 

The package was specifically designed for the analysis of *longitudinal* (e.g. multi-session) and/or *multi-site* neuroimaging data. 

Currently, `verywise` allows the estimation of vertex-wise **Linear Mixed Models**, but will be extended to other statistical models in the future.

It can handle imputed (phenotype) data from several packages (`mice`, `mi`, `amelia`, etc.).

Multiple testing correction is currently achieved using MCZ simulations from FreeSurfer.
This means that you will need FreeSurfer installed and correctly set up.


## Installation

You can install the development version of `verywise` from [GitHub](https://github.com/) with:

``` {r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("SereDef/verywise")
```

## Basic use

```{r load, eval = FALSE}
library(verywise)
```

```{r lmm, eval = FALSE}
# Run a linear mixed model
run_vw_lmm(
  formula = vw_thickness ~ sex * age + site + (1 | id), # model formula
  subj_dir = "./VeryWiseUserUser/SimulatedExample", # Neuro-imaging data location
  hemi = "both", # (default) which hemispheres to run
  pheno = phenotype, # An R object already in memory or path to file 
)

```
