---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# verywise: vertex-wise analysis of neuroimaging data

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of **`verywise`** is to offer a flexible, user-friendly interface to whole-brain (hemisphere) analysis of FreeSurfer MGH format data. 

The package allows application of several statistical models per vertex, including 
linear mixed models for the analysis of longitudinal and multi-site neuroimaging data.

It can handle imputed data from several packages (`mice`, `mi`, `amelia`, etc.).

Multiple testing correction is currently achieved using MCZ simulations from FreeSurfer.

## Installation

You can install the development version of `verywise` from [GitHub](https://github.com/) with:

``` {r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("SereDef/verywise")
```

```{r load, eval = FALSE}
# Load it up
library(verywise)
```

## Preparing your dataset 

You need FreeSurfer installed and a phenotype file that follows verywise structure. 

## Data poor? Simulate yourself some data, son
```{r simualte, eval = FALSE}

# Simulate FreeSurfer and phenotype dataset
simulate_data(subj_dir = "./VeryWiseUserUser/SimulatedExample")

```

## Basic use

```{r lmm, eval = FALSE}

# Run linear mixed model
run_vw_lmm(vw_thickness ~ sex * age + site + (1|id), # model formula
           subj_dir = "./VeryWiseUserUser/SimulatedExample", # Data location
           hemi = "lh", # left hemisphere
           n_cores = 1 # Number of cores for parallel processing
           )

```
