---
title: "Installation and system requirements"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

You can install the development version of `verywise` from [GitHub](https://github.com/). Note that you may need to install the [`devtools` package](https://devtools.r-lib.org/) first (if it is not already installed).


``` {r install, eval = FALSE}
# Install devtools (if not there yet)
if (!requireNamespace("devtools"))
  install.packages('devtools')

# Install verywise
devtools::install_github("SereDef/verywise")
```

This will also download some other R packages that are needed, so give it a minute. 
Finally, you can attach the package to make it easier to use, like so:

```{r setup}
library(verywise)
```

### System and software requirements
In order for `verywise` to run smoothly, you will need:
 - **`R`** (> 4.1) installed (dah).
 - A **Unix system**: `verywise` works on Mac and on most Linux distributions, but *not on Windows*. This is because we currently rely on functions from the [FreeSurfer suite](https://surfer.nmr.mgh.harvard.edu/) which require Unix/Mac. Getting `verywise` to work on Windows is not a strong priority for us.
 - [**Freesurfer**](https://surfer.nmr.mgh.harvard.edu/) installed.
