---
title: "Getting ready, step 1: Installation and system requirements"
author: "Serena Defina"
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

You can install the development version of `verywise` from [GitHub](https://github.com/). 

<ins>Note</ins>: you may need to install the [`devtools` package](https://devtools.r-lib.org/) first (if it is not already installed).

```{r install, eval = FALSE}
# Install devtools (if not there yet)
if (!requireNamespace("devtools"))
  install.packages('devtools')

# Install verywise
devtools::install_github("SereDef/verywise")
```

This will also download some other R packages that are needed, so give it a little minute. 

Finally, you can attach the package to make it easier to use, like so:

```{r setup}
library(verywise)
```

## System and software requirements

In order for `verywise` to run smoothly, you will need:

-   **`R`** (version \>= **4.1**) installed (*Dah!*).
-   A **Unix system**: `verywise` works on Mac and on most Linux distributions, but <ins>*not on Windows*</ins>. This is because we currently rely on functions from the FreeSurfer suite which require Unix/Mac.
-   [**FreeSurfer**](https://surfer.nmr.mgh.harvard.edu/) installed.

### Installing and setting up FreeSurfer

TODO
