---
title: "Optional: simulating data"
author: "Serena Defina"
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Data poor? Simulate yourself some data, son

If you are eager to get started with `verywise` but do not have access to a dataset (yet), 
you can generate both a set of ready-made FreeSurfer surface data files and a phenotype file to go with it. 

The simulated dataset will already be saved into a `verywise` structure. 

```{r setup}
library(verywise)
```

## Only generate the brain data 
If you have a phenotype dataset already, but you are missing some FreeSurfer data, you can call the following function:

```{r simualte brain, eval = FALSE}
# Simulate FreeSurfer dataset
my_dataset_structure = list("cohort1" = list("sessions" = c("01", "02"),
                                             "n_subjects" = 100),
                            "cohort2" = list("sessions" = c("01", "02"),
                                             "n_subjects" = 150)
                            )
my_random_seed = 1984

simulate_freesurfer_data(path = "./VeryWiseUser/SimulatedExample",
                         data_structure = my_dataset_structure,
                         vw_resolution = 163842,
                         measure = "thickness",
                         hemi = "lh",
                         fwhmc = "fwhm10",
                         target = "fsaverage",
                         vw_mean = 6.5,
                         vw_sd = 0.5,
                         simulate_association = NULL,
                         seed = my_random_seed)
```
In this example, we are going to generate 500 left hemisphere (`"lh"`) thickness maps. These maps belong to 250 fictional subjects, who all underwent 2 MRI sessions (`"01"` and `"02"`) and belong to two cohorts (`"cohort1"` and `"cohort2"`). The resolution of these fake surface maps is `163842` vertices (corresponding to the most detailed FreeSurfer template `"fsaverage"`), and they will have an overall mean of `6.5` (mm thickness) and a standard deviation of `0.5`.

You can also use the function to specify a simulated association with a variable of your choice.
TODO

## Only generate the phenotype data
On the other hand, if you already have some surfaces but you would like a phenotype to go with it you can use:
```{r simualte, eval = FALSE}
# Simulate phenotype dataset
simulate_long_pheno_data(data_structure = my_dataset_structure,
                         seed = my_random_seed)
```
This will generate a minimum "long format" dataframe with the variables you need, as well as simulated "sex" and "age" of the sample.

## Alright, all in one go now

```{r simualte, eval = FALSE}
# Simulate FreeSurfer and phenotype dataset
simulate_data(path = "./VeryWiseUser/SimulatedExample", 
              data_structure = my_dataset_structure,
              simulate_association =  0.05 * pheno$age,
              overwrite = TRUE,
              seed = my_random_seed)
```

This will call `simulate_long_pheno_data()` and `simulate_freesurfer_data()` to give you everything you need to play around with `verywise` model fitting.
