---
title: "Extracting and visualizing `verywise` results"
author: "Serena Defina"
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## `verywise` output

Check out the output directory. You should see something like this
[TODO: add image]

## Extracting mean/median cluster values 
Sometimes, you may want to extract the mean or the median value of a specific cluster from your results, for example to use this in further analysis. You can do this in `verywise` using the `significant_cluster_stats()` function. 

 <ins>Note</ins>: ideally, you should run your analyses with `save_ss = TRUE` or `save_ss = "path/to/ss"`, or you have called `build_supersubject()` in your pipeline, for this to sun smoothly.

```{r extract-cluster, eval = FALSE}
# Extract mean from significant clusters
df <- significant_cluster_stats(stat = "mean", # or "median"
                                ss_dir = "path/to/ss_directory", 
                                res_dir = "path/to/results", 
                                term = "age", # term of interest 
                                measure = "thickness", 
                                hemi = "lh")
```


## Visualizing results
To inspect and plot your results, you can use our interactive web application, [verywiseWIZard](https://github.com/SereDef/verywise-wizard). You can run this locally or try it out [here](https://seredef-verywise-wizard.share.connect.posit.cloud/).

Note: if you are using the online version of the WIZard, with results hosten on GitHub, you may want to look into the `move_result_files()` helper function, to organise your results in a way that is safe (i.e., does not expose individual level data) and quick.

Plots can also be generated using `verywise` like so:
```{r plot, eval = FALSE}
# Plot result brain map (requires FreeSurfer for templates and reticulate for interface with Python-based plotting libraries)
plot_vw_map(
  term = "age",
  hemi = "lh",
  measure = "area",
  res_dir = "/path/to/output",
  outline_rois = c("entorhinal", "precuneus"),
  fs_home = "/path/to/FREESURFER_HOME"
)
```

[TODO: expand with more details]

## On an HCP cluster 
If you want to make plots directly on an HPC cluster (e.g. Snellius) you will need to have some version of chrome installed (kaleido one works) and set up an Xvfb process **before** starting R (e.g. in your job script or in the shell before launching R):

For example, on Snellius, I do: 

```sh
module load 2025
module load Xvfb/21.1.18-GCCcore-14.2.0

Xvfb :99 -screen 0 1280x1024x24 &
export DISPLAY=:99
sleep 1  # give Xvfb time to start
```

Troubleshooting:

```sh
# check for existing Xvfb processed running on this node 
ps aux | grep -i "Xvfb.*:99"
```


## {-}
Next article: [Run vertex-wise federated / meta- analyses](06-run-vw-meta-and-fed.html)